Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data

### 2021-07-05 Manuscript/02-Method.Rmd
Finish revising method section.

### Spike-and-Slab Lasso Prior
We introduce a novel prior for GAMs, particularly for high-dimensional nonlinear effect modeling with bi-level selection. The proposed prior extends from the spike-and-slab lasso prior described in Equation \ref{eq:ssl}. Given the reparameterized deign matrix $\bs X_j = \begin{bmatrix} \bs X_j^0 : \bs X_j^\tp \end{bmatrix}^T$ for the $j$th variable, we impose a two-part SSL prior to the coefficients $\bs \beta_j$. Specifically, we impose independent group priors on the null space coefficients and on the penalized space coefficients respectively, 
$$
\begin{aligned}
  \beta^0_{j} |\gamma^0_{j},s_0,s_1 &\sim DE(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1)\\
  \beta^\tp_{jk} | \gamma^\tp_{j},s_0,s_1 &\sim DE(0,(1-\gamma^\tp_{j}) s_0 + \gamma^\tp_{j} s_1), 
\end{aligned}
$$
where $\gamma^0_{j}\in\{0,1\}$ and $\gamma^\tp_{j}\in \{0,1\}^{K_j}$ are latent indicator variables, indicating if the model includes the linear effect and the non-linear effect of the $j$th variable respectively. ; $s_0$ and $s_1$ are scale parameters, assuming $0 < s_0 < s_1$ and given. These scale parameters $s_0$ and $s_1$ can be treated as tuning parameters and optimized via cross-validation. A discussion of how to choose the scale parameters comes in Section \ref{selecting-optimal-scale-values}. To note, this prior differs from previous group spike-and-slab lasso priors[TODO: add Dr. Yi's citation], where the $\bs \beta^0_j$ and $\bs \beta^\tp_j$ shares the same indicator variables $\gamma_j^0$, $\gamma_j^\tp$ respectively. It is possible to relax the assumption such that each coefficient $\beta$ to have its own latent indicator $\gamma$, but at the cost ofcomplicating the bi-level function selection. Conversely, it is possible to add a more restrictive assumption on the priors, assuming that one indicator variable decides the inclusion of both the linear effect and non-linear effect, i.e. $\gamma_j = \gamma^0_j = \gamma^{pen}_j$. This converges^[Is "becomes" better?] to the SB-GAM [@Bai2021].

@@

The rest of the hierchical prior follows the traditional spike-and-slab lasso piror: we set up the hyper-prior of $\bs \gamma$, to allow local adaption of the shrinkage, using a bernoulli distribution. The two indicators of the $j$th predictor, $\gamma^{0}_j$ and $\gamma^{pen}_j$, shares the same probability parameter,
$$
\begin{aligned}
\gamma_{j}^{0} | \theta_j &\sim Bin(\gamma^{0}_{j}|1, \theta_j)\\
\gamma_{j}^{pen} | \theta_j &\sim Bin(\gamma^{pen}_{j}|1, \theta_j).
\end{aligned}
$$

This is to leverage the fact that the probability of selecting the bases of a smooth function should be similar, while allowing different penalty on the null space and penalty space of the spline function. The hyper prior of $\gamma_{j}^{0}$ decides the sparsity of the model at the function selection level, while that of $\gamma_{j}^{pen}$ decides the smoothness of the spline function at basis function level. Meanwhile, we specify that $\gamma_{j}^0$ and $\gamma_{j}^{pen}$ are independently distributed for analytic simplicity. We further specify the parameter $\theta_j$ follows a beta distribution with given shape parameters $a$ and $b$,
$$
\theta_j \sim Beta(a, b).
$$
The beta distribution is a conjugate prior for the binomial distribution and hence provides some computation convenience. For simplicity, we focus on a special case of beta distribution, uniform (0,1). When the variable have large effects in any of the bases, the parameter $\theta_j$ will be estimated large, which in turn encourages the model to include the rest of bases, achieving the local adaption among spline bases. Hereafter, we refer Bayesian hierarchical generalized additive models with the spike-and-slab lasso prior as the BHAM-SSL.

\newpage

```{r}
include_graphics("Fig/SS_prior_flow_chart.png")
```

@@

Considering the dependency of spline bases, the grouping nature of the bases should be taken care of. Naturally, the Bayesian hierarchical models allow us easily to take account of the dependency of the spline bases as grouping information and provide more reliable estimation. Our hierarchical GAMs specify the spike-and-slab mixture normal prior on the coefficients: for the coefficients of the $j$th predictor, $\beta^0_j$ and $\beta^{pen}_{jk}$, we have
$$
\begin{aligned}
  \beta^0_{j} |\gamma^0_{j},s_0,s_1 &\sim N(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1)\\
  \beta^{pen}_{jk} | \gamma^{pen}_{j},s_0,s_1 &\sim N(0,(1-\gamma^{pen}_{j}) s_0 + \gamma^{pen}_{j} s_1), 
\end{aligned}
$$
where $\gamma^0_{j}\in\{0,1\}$ and $\gamma^{pen}_{j}\in \{0,1\}^{K_j}$ are latent indicator variables, indicating if the model includes the linear effect and non-linear effect of the $j$th variable respectively; $s_0$ and $s_1$ are scale parameters, assuming $0 < s_0 < s_1$ and given.  Thus, the priors are formulated as a mixture of the shrinkage prior normal(0, $s_0$) and the weakly informative prior normal(0, $s_1$), which are spike and slab components respectively. Like any other spike and slab priors, the spike density is to contain the minimum to zero effects, while the slab density is to allow large effects. The scale parameters $s_0$ and $s_1$ can be treated as tuning parameters, which can be optimized via cross-validation. A discussion of how to choose the scale parameters comes later.

To note here, unlike the original spike-and-slab prior where each predictor have only indicator variable, $\gamma_j$, we have two indicators associated with each predictor, $\gamma^0_j$ and $\gamma^{pen}_j$ to decide the inclusion of the linear effect and non-linear effect of the predictor respectively. It is possible to add more constraints on the model, assuming that one indicator variable decides the inclusion of both the linear effect and non-linear effect, i.e. $\gamma_j = \gamma^0_j = \gamma^{pen}_j$. Such formulations, commonly referred as "all-in-all-out", has been implemented in many high dimensional Bayesian GAMs, see Yang & Narisetty (2020) and Bai (Work in progress). The "all-in-all-out" set-up was criticized for its inability to answering an important research question, if the effects are non-linear or not. On the other hand, the model can be much more flexible by allowing each dimension of the penalized space to have an independent indicator, i.e. $\gamma^{pen}_{jk}$ of $\beta^{pen}_{jk}$ for $k = 1\dots, K_j$. This prior set-up generalized to the group spike-and-slab. The additional flexibility doesn't not necessarily grand the model neither theoretical justification nor analytic simplicity. The spline penalty assumes a common smooth shrinkage $\lambda_j$ applied to all coefficients of the penalized space, which contradicts to the model setup. Besides, having multiple indicator variables complicates the decision of if non-linear effects are necessary.

We set up the hyper-prior of $\bs \gamma$ to allow local adaption of the shrinkage, using a binomial distribution. The two indicators of the $j$th predictor, $\gamma^{0}_j$ and $\gamma^{pen}_j$, shares the same probability parameter,
$$
\begin{aligned}
\gamma_{j}^{0} | \theta_j &\sim Bin(\gamma^{0}_{j}|1, \theta_j)\\
\gamma_{j}^{pen} | \theta_j &\sim Bin(\gamma^{pen}_{j}|1, \theta_j).
\end{aligned}
$$

This is to leverage the fact that the probability of selecting the bases of a smooth function should be similar, while allowing different penalty on the null space and penalty space of the spline function. The hyper prior of $\gamma_{j}^{0}$ decides the sparsity of the model at the function selection level, while that of $\gamma_{j}^{pen}$ decides the smoothness of the spline function at individual predictor level. Meanwhile, we specify that $\gamma_{j}^0$ and $\gamma_{j}^{pen}$ are independently distributed for analytic simplicity. However, this set-up could possibility result in a situation that is not theoretically possible: the non-linear component is selected, but the linear component is not. This can be addressed analytically by forcing to include the linear component when non-linear component is included. Another possible solution is to impose an dependent structure of $\gamma_{j}^{pen}$ on $\gamma_{j^{0}}$, i.e. $\gamma_j^{pen}|\gamma_{j}^{0}, \theta_j$.

 <!-- Such prior set-ups ignore the null space of the smooth function, and entirely removes the variable out of the function. The "all-in-all-out" assumption makes differentiate linear and non-linear effect harder. In contrasts, our prior set up provides a more flexible solution, as the penalty for smoothness is locally adapative for the bases and hence retain the the null space if necessary.  -->

We further specify the parameter $\theta_j$ follows a beta distribution with given shape parameters $a$ and $b$,
$$
\theta_j \sim Beta(a, b).
$$
 The beta distribution is a conjugate prior for the binomial distribution and hence provides some computation convenience. Specifically, we focus on a special case of beta distribution, uniform (0,1) for simplicity and convenience. To note, when the variable have large effects in any of the bases, the parameter $\theta_j$ will be estimated large, which in turn encourages the model to include the rest of bases, achieving the local adaption among spline bases. Hereafter, we refer the hierarchical spline GAM with the spike-and-slab mixture normal prior as the ssGAM-MN.

One of the critics received by the spike-and-slab mixture normal prior is that the tails of a normal distribution diminishes to zero too fast, which renders problem when estimating the large effects. In contrast, distributions with heavier tails are preferred, for example $t$ distribution with small degree of freedom and double exponential distribution. Specifically, double exponential prior is equivalent to the LASSO penalty in the frequentist models, providing a sparse solution.

@@

To alleviate the computation burden, many methodology development studies focus on the maximum a posteriori (MAP) estimate [@Tipping2004]. EM-based algorithms are prevalent used to speed the computation. @Rockova2014a proposed an EM-based algorithm for a spike-and-slab Gaussian mixture prior, EMVS. The coefficient $\beta_j$ independently follows a mixture normal distribution $N(0, \sigma^2((1-\gamma_i)v_0 + \gamma_i v_1)$ for $0 \leq v_0 < v_1$. The parameter $\sigma^2$ follows a inverse gamma prior, $IG(v/2, v\lambda /2)$. Like in the spike-and slab family, the latent binary variables $\bs \gamma$ follows a Bernoulli prior with a hyper prior $\theta$ follows a beta distribution, or simpler the uniform (0,1). In the E step, the latent variable are treated as the missing data, and calculate the posterior mean of $\gamma$; in the M step, a ridge estimator was used to update the coefficients, and $\sigma^2, \theta$ are updated accordingly. As the ridge regression creates small but nonzero coefficients, a further step is needed to select the variable: EMVS further suggests to threshold the inclusion probability for variable selection or generate regularization plot with different values of the spike scale parameter. EMVS is also compatible with structured prior when dealing with grouped variables. The group structure can be imposed on the prior of the latent binary variable $\gamma$ via either logistic regression or markov random field.

The same authors [@Rockova2018b; @Rockova2018] further extended the continuous spike-and-slab Gaussian prior to the spike-and-slab double exponential distribution, as known as spike-and-slab lasso (SSL). The authors also lay the theoretical discussion of comparing spike-and-slab from the penalty perspective and the Bayesian perspective. The SSL also mitigated the problem of EMVS that the weak signals are not shrink to zero. Yi and his group incorporated the SSL prior in the EM-IWLS algorithm and proposed another EM algorithm, EM-cyclic coordinate descent, based on the Lasso penalty SSL exhibit. Such application were used in generalized linear models[@Tang2017a], cox models [@Tang2017] and their grouped counterparts[@Tang2018; @Tang2019]. Both EM based algorithm provide deterministic solutions, which becomes a more popular property as reproducible research becomes more valued. However, while the EMVS are not able to provide inference of the coefficients easily, the EM-IWLS can easily extract the variance-covariance matrix due to the IWLS algorithm. @Bai2020 provides an up-to-date summary of SSL methods and theoretical discussion.

 Given the proposed model specification, the estimation of the parameters can be easily achieved with MCMC algorithm. Nevertheless, considering the computational burden MCMC algorithms bear especially in high-dimensional analysis, we propose parsimonious solutions, stemming from previous EM-IWLS [@Rockova2014a]and EM-coordinate descent algorithms[@Tang2018; @Tang2019]. The computational friendliness allows for fitting a model with different sets of initial values, mimicking the posterior sampling in the MCMC algorithms. We are hopeful that this could improve the sacrificed robustness of the estimation process due to the MAP estimation used by the EM algorithms in comparison to the model averaging approach.

Parsimonious computation is always encouraged in high-dimension data analysis. Normally, Bayesian methodology loses its advantages over Frequentest penalized model because of the computation cost. Previous Bayesian spline models heavily relies on the MCMC algorithm to establish posterior distribution of parameters. One of the exception is Bai (in progress), who a grouped spike-and-slab LASSO prior [@Rockova2018] to address the grouping structure of spline bases. He focused on the MAP estimator of the parameters, but failed to address the uncertainty measure of the estimates. The main difference between Bai's work and the proposed models are two-folded: 1) the proposed models are feasible to answer the question if the non-linear effects are necessary for the predictors, while Bai's work can't due to the "all-in-all-out" approach; 2) the proposed models can provide uncertainly measures in addition to the point estimates of the coefficients, while Bai's work can't not due to the model fitting algorithm.

In the follow section, we introduced two expectation-maximization (EM) based algorithms for fast computing, EM-IRLS and EM-coordinate Descent algorithm. Instead of simulating the posterior distribution via MCMC, we iteratively estimate the MAP. Compared to the EM-IRLS algorithm, the EM-coordinate descent provides an expedited model fitting process at the cost of model uncertainty.

@@

For ssGAM, we define the parameters of interest as $\Theta = {\bs \beta, \bs \theta, \phi}$ and consider the latent binary indicators $\bs \gamma$ as nuisance parameters of the model.
Our objective is to find the parameters $\Theta$ that maximize the posterior density function,or equivalently, the logarithm of the density function, 
$$
\begin{aligned}
& \text{argmax}_{\Theta}
\log p(\Theta, \bs \gamma| \textbf{y}, \textbf{X}) \\
&= \log p(\textbf{y}|\bs \beta, \phi) + \log p(\phi) + \sum\limits_{j=1}^p\left[\log p(\beta^0_j|\gamma^0_j)+\sum\limits_{k=1}^{K_j} \log p(\beta^{pen}_{jk}|\gamma^{pen}_{jk})\right]\\
& +\sum\limits_{j=1}^{p} \left[ (\gamma^0_j+\gamma_{j}^{pen})\log \theta_j + (2-\gamma^0_j-\gamma_{j}^{pen}) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log p(\theta_j)
\end{aligned}
$$  
<!-- For simplicity, we consider the null space to be 1-dimension, and rewrite $\beta^0_j$ as $\beta_{j0}$, and hence removing the superscript $^{pen}$ from the coefficients of the the penalized space. Using the new notation, the part component $\log p(\beta^0_j|\gamma^0_j)+\sum\limits_{k=1}^{K_j} \log p(\beta^{pen}_{jk}|\gamma^{pen}_{jk})$ can simplifies to $\sum\limits_{k=0}^{K_j} \log p(\beta_{jk}|\gamma_{jk})$. Without loos of generality, the model still work for the spline function whose dimensionality of the null space is more than 1. -->

We use the EM algorithm to find the MAP estimate of $\Theta$. Since the latent binary indicators $\bs \gamma$ are not of our primary interest, we treat them as as the "missing data" in the EM algorithms. Naturally In the E-step, we calculate the expectation of posterior density function of $\log p(\Theta, \bs \gamma| \textbf{y}, \textbf{X})$ with respect to the latent indicators $\bs \gamma$ conditioning on the values from previous iteration $\Theta^{t-1}$, 
$$
E_{\bs \gamma|\Theta^{t-1}}\log p(\Theta, \bs \gamma| \textbf{y}, \textbf{X}) .
$$
Hereafter, we use the short hand notation $E(\cdot)\equiv E_{\bs \gamma|\Theta^{t-1}}(\cdot)$. In the M-step, we find the $\Theta^{t}$ that maximize $E_{\bs \gamma|\Theta^{t-1}}\log p(\Theta, \bs \gamma| \textbf{y}, \textbf{X})$. To note here, the log-posterior of ssGAM (up to additive constants) can be written as a two-part equation

$$ \log p(\Theta, \bs \gamma| \textbf{y}, \textbf{X}) = Q_1(\beta, \phi) + Q_2 (\gamma,\theta),$$
Where
$$ Q_1(\bs \beta, \phi) = \log p(\textbf{y}|\bs \beta, \phi) + \log p(\phi) + \sum\limits_{j=1}^p\left[\log p(\beta^0_j|\gamma^0_j)+\sum\limits_{k=1}^{K_j} \log p(\beta^{pen}_{jk}|\gamma^{pen}_{jk})\right]$$
and $$
Q_2(\bs\gamma,\bs\theta) = \sum\limits_{j=1}^{p} \left[ (\gamma^0_j+\gamma_{j}^{pen})\log \theta_j + (2-\gamma^0_j-\gamma_{j}^{pen}) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log p(\theta_j) .$$
$Q_1$ and $Q_2$ are respectively, the log-likelihood of model coefficients and the dispersion parameter $\bs \beta, \phi$, and latent indicators $\bs \gamma, \theta$. Conditioning on $\gamma_{jk}$, $Q_1$ and $Q_2$ are independent and can be maximized separately for $\bs \beta, \phi$ and $\bs \theta$.

The E- and M- steps are iterated until the algorithm converge. Depending on the choice of the mixture priors, either mixture normal or mixture double exponential, the maximization in the M-step can take different algorithms, IRLS for uncertainty measures and coordinate descent for faster computation.

\

The main idea of the EM-IRLS algorithm is to use the iterative re-weighted least square algorithm to find the estimate $\bs \beta, \phi$ that maximizes $E(Q_1)$. Recall the prior set-up of mixture normal prior, we have the prior densities of $\bs \beta$ conditioning on $\bs \gamma$

$$
\begin{aligned}
p(\bs \beta | \bs\gamma) &\propto \prod\limits_{j=1}^{p}\left[{S^0_j}^{-1/2}\exp(-1/2({\beta^0_{j}}^2/S^0_{j}))\prod\limits_{k=1}^{K_j}S_{j}^{-1/2}\exp(-1/2(\beta_{jk}^2/S_{j}))\right],%\\
% p(\bs\gamma | \bs \theta) & = \prod\limits_{j=1}^{p}\prod\limits_{k=1}^{K_j} \theta_j^{\gamma_{jk}} (1-\theta_j)^{1-\gamma_{jk}}.
\end{aligned}
$$
and 
\begin{align*}
E(Q_1) &= \log p(\textbf{y}|\bs \beta, \phi) + \log p(\phi) + \sum\limits_{j=1}^{p} \left[E({S^0_j}^{-1}){\beta^0_j}^2+\sum\limits_{k=1}^{K_j}E(S^{-1}_{j})\beta_{jk}^2\right]\\
E(Q_2) &= \sum\limits_{j=1}^{p}\sum\limits_{k=1}^{K_j} \left[ E(\gamma_{jk})\log \theta_j + (1-E(\gamma_{jk})) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log p(\theta_j)
\end{align*}
where $S_{j}^0 = (1-\gamma^{0}_{j}) s_0 + \gamma^{0}_{j} s_1$ and $S_{j} = (1-\gamma^{pen}_{j}) s_0 + \gamma^{pen}_{j} s_1$. To calculate two unknown quantities $E({S_j^0}^{-1})$ and $E(S^{-1}_j)$, the posterior probability $p^0_{j} \equiv p(\gamma^{0}_{j}=1|\Theta^{t-1})$ and $p_{j} \equiv p(\gamma^{pen}_{j}=1|\Theta^{t-1})$ are necessary, which can be derived via Bayes' theorem. The calculation for $p_j$ is slightly different from that of $p^0_j$, as $p_j$ depends on the value of $\beta_{jk}$ for $k=1, \dots, K_j$ and $p^0_j$ only depends on $\beta_j^0$. The calculation follows the equations below.

and $Pr(\gamma_{j}^{0} = 1|\theta_j) =Pr(\gamma_{j}^{pen} = 1|\theta_j) = \theta_j$, $Pr(\gamma_{j}^{0} = 0|\theta_j) = Pr(\gamma_{j}^{pen} = 0|\theta_j) = 1-\theta_j$, $f(\beta|\gamma=1, s_1) = \text{Normal}(\beta|0 , s_1)$, $f(\beta|\gamma=0, s_0) = \text{Normal}(\beta|0 , s_0)$. It is trivial to show
$$
\begin{aligned}
&E(\gamma^0_{j})  = p^0_{j} & &E(\gamma^{pen}_{j})  = p_{j}\\
&E({S^0}^{-1}_{j}) = \frac{1-p^{0}_{j}}{s_0} + \frac{p^{0}_{j}}{s_1} & &E(S^{-1}_{j}) = \frac{1-p_{j}}{s_0} + \frac{p_{j}}{s_1}.
\end{aligned}
$$

>>>>

### Spike-and-Slab Lasso Prior
We introduce a novel prior for GAMs, particularly for high-dimensional nonlinear effect modeling with bi-level selection. The proposed prior extends from the spike-and-slab lasso prior described in Equation \ref{eq:ssl}. Given the reparameterized deign matrix $\bs X_j = \begin{bmatrix} \bs X_j^0 : \bs X_j^\tp \end{bmatrix}^T$ for the $j$th variable, we impose a two-part SSL prior to the coefficients $\bs \beta_j$. Specifically, we impose independent group priors on the null space coefficients and on the penalized space coefficients respectively,

\begin{align}\label{eq:bham_ssl}
  \beta^0_{j} |\gamma^0_{j},s_0,s_1 &\sim DE(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1)\\
  \beta^\tp_{jk} | \gamma^\tp_{j},s_0,s_1 &\sim DE(0,(1-\gamma^\tp_{j}) s_0 + \gamma^\tp_{j} s_1), 
\end{align}

where $\gamma^0_{j}\in\{0,1\}$ and $\gamma^\tp_{j}\in \{0,1\}^{K_j}$ are latent indicator variables, indicating if the model includes the linear effect and the non-linear effect of the $j$th variable respectively. ; $s_0$ and $s_1$ are scale parameters, assuming $0 < s_0 < s_1$ and given. These scale parameters $s_0$ and $s_1$ can be treated as tuning parameters and optimized via cross-validation. A discussion of how to choose the scale parameters comes in Section \ref{selecting-optimal-scale-values}. To note, this prior differs from previous group spike-and-slab lasso priors[TODO: add Dr. Yi's citation], where the $\bs \beta^0_j$ and $\bs \beta^\tp_j$ shares the same indicator variables $\gamma_j^0$, $\gamma_j^\tp$ respectively. It is possible to relax the assumption such that each coefficient $\beta$ to have its own latent indicator $\gamma$, but at the cost ofcomplicating the bi-level function selection. Conversely, it is possible to add a more restrictive assumption on the priors, assuming that one indicator variable decides the inclusion of both the linear effect and non-linear effect, i.e. $\gamma_j = \gamma^0_j = \gamma^\tp_j$. This converges^[Is "becomes" better?] to the SB-GAM [@Bai2021].

@@

The rest of the hierchical prior follows the traditional spike-and-slab lasso piror: we set up the hyper-prior of $\bs \gamma$, to allow local adaption of the shrinkage, using a bernoulli distribution. The two indicators of the $j$th predictor, $\gamma^{0}_j$ and $\gamma^\tp_j$, shares the same probability parameter,
$$
\begin{aligned}
\gamma_{j}^{0} | \theta_j &\sim Bin(\gamma^{0}_{j}|1, \theta_j)\\
\gamma_{j}^\tp | \theta_j &\sim Bin(\gamma^\tp_{j}|1, \theta_j).
\end{aligned}
$$

This is to leverage the fact that the probability of selecting the bases of a smooth function should be similar, while allowing different penalty on the null space and penalty space of the spline function. The hyper prior of $\gamma_{j}^{0}$ decides the sparsity of the model at the function selection level, while that of $\gamma_{j}^\tp$ decides the smoothness of the spline function at basis function level. Meanwhile, we specify that $\gamma_{j}^0$ and $\gamma_{j}^\tp$ are independently distributed for analytic simplicity. We further specify the parameter $\theta_j$ follows a beta distribution with given shape parameters $a$ and $b$,
$$
\theta_j \sim Beta(a, b).
$$
The beta distribution is a conjugate prior for the binomial distribution and hence provides some computation convenience. For simplicity, we focus on a special case of beta distribution, uniform (0,1). When the variable have large effects in any of the bases, the parameter $\theta_j$ will be estimated large, which in turn encourages the model to include the rest of bases, achieving the local adaption among spline bases. Hereafter, we refer Bayesian hierarchical generalized additive models with the spike-and-slab lasso prior as the BHAM-SSL.

```{r fig.height=4, fig.width=6}
# TODO: Refactor the graph to a latex flowchart
include_graphics("Fig/SS_prior_flow_chart.png")
```

@@

With the re-parameterization step of the basis function matrix $X$, it is possible to generalized the SSL prior to other priors, for example normal priors for ridge-type regularizations and mixture normal prior for spike-and-slab
regularization. These priors would work better in low and medium dimensional settings where the sparse assumption is not necessary. Here we focus on the mixture normal prior as a demonstration of applying continous spike-and-slab prior in additive models. ^[TODO: enter a sentence introducing EMVS and spike-and-slab mixture normal prior].

A spike-and-slab mixture normal prior in additive models can be expressed as 
$$
\begin{aligned}
  \beta^0_{j} |\gamma^0_{j},s_0,s_1 &\sim N(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1)\nonumber\\
  \beta^\tp_{jk} | \gamma^\tp_{j},s_0,s_1 &\sim N(0,(1-\gamma^\tp_{j}) s_0 + \gamma^\tp_{j} s_1).
\end{aligned}
$$
Similar to the BHAM-SSL prior in Equation \ref{eq:bham_ssl}, $0 < s_0 < s_1$ are tuning parameters and can be optimized via cross-validation. One of the critics received by the spike-and-slab mixture normal prior is that the tails of a normal distribution diminishes to zero too fast, which renders problem when estimating the large effects. Distributions with heavier tails can be used as an alternative, for example mixture $t$ distribution with small degree of freedom.

@@

Given the proposed model specifications, these models can be easily fitted with MCMC algorithms. Nevertheless,the computational burden MCMC algorithms bear creates scalability issues. George and McCulloch[@George1997] examined the computation speed for various MCMC algorithms with continous spike-and-slab priors, and suggested MCMC algorithms works well for medium size of predictors ($p$=25). However, it is not feasible to high-dimensionality data analysis where the number of predictors easily exceeds 100. Sepcific to additive models, each predictor would expand to mutliple new "predictors" via basis functions, creating greater computational demands. @Scheipl2013 demonstrated the computational demands of MCMC algorithm for fitting spike-and-slab additive models grow exponentially as $p$ increases modestly via monte carlo simulation studies. Hence, we feel compelled to develop scalable and parsimonious algorithms for fitting Bayesian hierarchical additive models in high-dimensional settings.

Alternative to the sampling algorithms for fitting Bayesian models, optimization algorithms focus on the maximum a posteriori (MAP) estimates and speed up the process at the cost of uncertainty measures. The earliest work for fitting spike-and-slab models using optimization algorithm is EMVS [@Rockova2014a]. @Rockova2014a proposed an expectation-maximization (EM) based algorithm to fit models that use continous mixture normal priors. 
<!-- The coefficient $\beta_j$ independently follows a mixture normal distribution $N(0, \sigma^2((1-\gamma_i)v_0 + \gamma_i v_1)$ for $0 \leq v_0 < v_1$. The parameter $\sigma^2$ follows a inverse gamma prior, $IG(v/2, v\lambda /2)$. Like in the spike-and slab family, the latent binary variables $\bs \gamma$ follows a Bernoulli prior with a hyper prior $\theta$ follows a beta distribution, or simpler the uniform (0,1).  -->
In the E step, the latent variables $\gamma$ are treated as the missing data, and calculate the posterior mean of the density function; in the M step, a ridge estimator was used to update the coefficients, and $\sigma^2, \theta$ are updated accordingly. 
<!-- As the ridge regression creates small but nonzero coefficients, a further step is needed to select the variable: EMVS further suggests to threshold the inclusion probability for variable selection or generate regularization plot with different values of the spike scale parameter. EMVS is also compatible with structured prior when dealing with grouped variables. The group structure can be imposed on the prior of the latent binary variable $\gamma$ via either logistic regression or markov random field.  -->
The same authors [@Rockova2018b; @Rockova2018] further combined the EM algorithm with coodirnate descent (CD) algoirthm for fitting SSL models.
<!-- the continuous spike-and-slab Gaussian prior to the spike-and-slab double exponential distribution, as known as spike-and-slab lasso (SSL). The authors also lay the theoretical discussion of comparing spike-and-slab from the penalty perspective and the Bayesian perspective. The SSL also mitigated the problem of EMVS that the weak signals are not shrink to zero. -->
Yi and his group incorporated the SSL prior in the EM-Iterative Weighted Least Square(IWLS) algorithm and proposed another EM algorithm, EM-cyclic coordinate descent, based on the Lasso penalty SSL exhibit. Such applications were used in generalized linear models[@Tang2017a], cox models [@Tang2017] and their grouped counterparts[@Tang2018; @Tang2019]. Both EM based algorithms provide deterministic solutions, which becomes a more popular property as reproducible results.  @Bai2020 provides an up-to-date summary of SSL methods and theoretical discussion.

In the follow section, we expand the two EM-based algorithms, EM-CD and EM-IWLS algorithms, to fit BHAMs. To note, the two proposed algorithm provides different utilities. The EM-CD algorithm is specifically for fitting BHAM-SSL with an expediated performance, recommending to use in high and ultra-high dimensional setting. SB-GAM[@Bai2020; @Bai2021] also used EM-CD algorithm. The main difference between proposed EM-CD algorithm and SB-GAM is that SB-GAM uses a blocked CD algorithm for their group prior. In comparison, our prior is pairwise independent requiring no speical treatment. A sepcifica concern of EM-CD algorithm is that it provides no information for inference.  In the contrast, the EM-IWLS provides a complement to the problem. The EM-IWLS can easily extract the variance-covariance matrix due to the IWLS algorithm, and used to construct Wald type inference. Moreover, the EM-IWLS is a more general model fitting algorithm that can be used for fitting not only SSL and continuous SS pirors but also bayesian ridge and lasso priors.

<!-- , stemming from previous EM-coordinate descent algorithms[@Tang2018; @Tang2019] and EM-IWLS [@Rockova2014a]. Instead of simulating the posterior distribution via MCMC, we iteratively estimate the MAP. Compared to the EM-IRLS algorithm, the EM-coordinate descent provides an expedited model fitting process at the cost of model uncertainty.  -->

<!-- The main difference between Bai's work and the proposed models are two-folded: 1) the proposed models are feasible to answer the question if the non-linear effects are necessary for the predictors, while Bai's work can't due to the "all-in-all-out" approach; 2) the proposed models can provide uncertainly measures in addition to the point estimates of the coefficients, while Bai's work can't not due to the model fitting algorithm. One of the exception is Bai (in progress), who a grouped spike-and-slab LASSO prior [@Rockova2018] to address the grouping structure of spline bases.  -->

@@

For BHAMs, we define the parameters of interest as $\Theta = {\bs \beta, \bs \theta, \phi}$ and consider the latent binary indicators $\bs \gamma$ as nuisance parameters of the model. Our objective is to find the parameters $\Theta$ that maximize the posterior density function, or equivalently, the logarithm of the density function, 
$$
\begin{aligned}
& \text{argmax}_{\Theta}
\log p(\Theta, \bs \gamma| \textbf{y}, \textbf{X}) \\
&= \log p(\textbf{y}|\bs \beta, \phi) + \log p(\phi) + \sum\limits_{j=1}^p\left[\log p(\beta^0_j|\gamma^0_j)+\sum\limits_{k=1}^{K_j} \log p(\beta^{pen}_{jk}|\gamma^{pen}_{jk})\right]\\
& +\sum\limits_{j=1}^{p} \left[ (\gamma^0_j+\gamma_{j}^{pen})\log \theta_j + (2-\gamma^0_j-\gamma_{j}^{pen}) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log p(\theta_j)
\end{aligned}
$$  
For simplicity, we consider the null space to be 1-dimension.

<!-- , and rewrite $\beta^0_j$ as $\beta_{j0}$, and hence removing the superscript $^{pen}$ from the coefficients of the the penalized space. Using the new notation, the part component $\log p(\beta^0_j|\gamma^0_j)+\sum\limits_{k=1}^{K_j} \log p(\beta^{pen}_{jk}|\gamma^{pen}_{jk})$ can simplifies to $\sum\limits_{k=0}^{K_j} \log p(\beta_{jk}|\gamma_{jk})$. Without loos of generality, the model still work for the spline function whose dimensionality of the null space is more than 1. -->

We use the EM algorithm to find the MAP estimate of $\Theta$. Since the latent binary indicators $\bs \gamma$ are not of our primary interest, we treat them as the "missing data" in the EM algorithms. Naturally In the E-step, we calculate the expectation of posterior density function of $\log p(\Theta, \bs \gamma| \textbf{y}, \textbf{X})$ with respect to the latent indicators $\bs \gamma$ conditioning on the values from previous iteration $\Theta^{t-1}$, 
$$
E_{\bs \gamma|\Theta^{t-1}}\log p(\Theta, \bs \gamma| \textbf{y}, \textbf{X}) .
$$
Hereafter, we use the short hand notation $E(\cdot)\equiv E_{\bs \gamma|\Theta^{t-1}}(\cdot)$. In the M-step, we find the $\Theta^{t}$ that maximize $E\log p(\Theta, \bs \gamma| \textbf{y}, \textbf{X})$. To note here, the log-posterior density of BHAMs (up to additive constants) can be written as a two-part equation

$$ \log p(\Theta, \bs \gamma| \textbf{y}, \textbf{X}) = Q_1(\beta, \phi) + Q_2 (\gamma,\theta),$$
Where
$$ Q_1(\bs \beta, \phi) = \log p(\textbf{y}|\bs \beta, \phi) + \log p(\phi) + \sum\limits_{j=1}^p\left[\log p(\beta^0_j|\gamma^0_j)+\sum\limits_{k=1}^{K_j} \log p(\beta^{pen}_{jk}|\gamma^{pen}_{jk})\right]$$
and $$
Q_2(\bs\gamma,\bs\theta) = \sum\limits_{j=1}^{p} \left[ (\gamma^0_j+\gamma_{j}^{pen})\log \theta_j + (2-\gamma^0_j-\gamma_{j}^{pen}) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log p(\theta_j) .$$
$Q_1$ and $Q_2$ are, respectively, the log-likelihoods of model coefficients and the dispersion parameter $\bs \beta, \phi$, and latent indicators $\bs \gamma, \theta$. Conditioning on $\gamma_{jk}$, $Q_1$ and $Q_2$ are independent and can be maximized separately for $\bs \beta, \phi$ and $\bs \theta$.

The E- and M- steps are iterated until the algorithm converge. Depending on the choice of the mixture priors, the maximization in the M-step can take different algorithms, CD for faster computation and IWLS for uncertainty measures.
